---
title: "Stat 795 Check-in 2"
author: "Heidi Rogers"
date: "2024-02-05"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Part 1: Bootstrapping

```{r, warning=FALSE, message=FALSE}
## PROBLEM 1

boot1 <- read.csv("Boot1.csv")

set.seed(100)
bootstrap1 <- c()
nsamples <- 1000

for(i in 1:nsamples){
  # sample from boot1
  current_sample <- sample(boot1$GammaData, replace=TRUE)
  # calculate standard deviation of the sample and add it to bootstrap vector
  bootstrap1[i] <- sd(current_sample)
}

# 95% confidence interval for population standard deviation
quantile(bootstrap1, probs=c(.025,.975))
```

```{r, warning=FALSE, message=FALSE}
## PROBLEM 2

boot2 <- read.csv("Boot2.csv")

set.seed(200)
bootstrap2 <- c()
nsamples <- 1000

for(i in 1:nsamples){
  # sample from boot2
  current_sample <- sample(boot2$Data, replace=TRUE)
  # calculate IQR of the sample and add it to bootstrap vector
  q3 <- quantile(current_sample, probs = .75) # 75% quartile
  q1 <- quantile(current_sample, probs = .25) # 25% quartile
  bootstrap2[i] <- q3 - q1 
}

# 90% confidence interval for population IQR
quantile(bootstrap2, probs=c(.05,.95))
```

```{r, warning=FALSE, message=FALSE}
## PROBLEM 3

boot3 <- read.csv("Boot3.csv")

set.seed(300)
bootstrap3 <- c()
nsamples <- 1000

for(i in 1:nsamples){
  # sample from boot3
  current_sample <- sample(boot3$Data, replace=TRUE)
  # calculate median of the sample and add it to bootstrap vector
  bootstrap3[i] <- median(current_sample)
}

# 99% confidence interval for population median
quantile(bootstrap3, probs=c(.005,.995))
```

## Part 2: Permutation Tests

```{r, warning=FALSE, message=FALSE}
## PROBLEM 1

perm1 <- read.csv("PermTest1.csv")
set.seed(1000)

# observed difference in sample medians
med_obs <- median(perm1$Group1) - median(perm1$Group2) # -0.3

# specify lengths of groups
n1 <- length(perm1$Group1); n2 <- length(perm1$Group2)
N <- n1 + n2 # total observations

#comibine data
alldata <- c(perm1$Group1, perm1$Group2)

# initialize variables
nperms = 1000
all_med_perm <- c()
p_val <- 0
count <- 0

# permute data and calculate difference in medians 1000 times
for(i in 1:nperms){ 
  # permute data
  index = sample(1:N, n1)
  data1 = alldata[index]
  data2 = alldata[-index]
  
  # calculate difference in medians
  med_perm <- median(data1) - median(data2)
  all_med_perm[i] <- med_perm
  
  # for two-tailed p-value:
  # count = number of times the absolute value of the sample difference 
  # is >= absolute value of the observed difference
  if(abs(med_perm) > abs(med_obs) | abs(med_perm) == abs(med_obs)){
    count <- count + 1
  }
  # calculate p-value
  p_val <- count / nperms
}

p_val #.341

# histrogram of permuted statistisc
ggplot() + 
  geom_histogram(aes(x=all_med_perm), bins = 15, color="black", fill="lightblue") +
  labs(x = "Difference in Median Between Groups", 
       title="Distribution of 1000 Permuted Statistics", 
       y = "Count") +
  geom_vline(xintercept = med_obs, 
             linetype = "solid", color = "red", size = .7) +
  geom_text(aes(x=-.4, label = "Observed", y=200), color="red", angle=45) +
  theme_minimal()

```

**Ho:** population medians are the same between groups (m1 - m2 = 0, where m1 = population median for group 1)

**Ha:** population medians are different between group 1 and group 2 (m1 - m2 =/ 0)

The observed difference in medians between the two groups is -0.3. The p-value for the 1000 permutations is .341 \> .05, therefore, we fail to reject the null and do not have enough evidence to say that the two groups have different population medians.

```{r, warning=FALSE, message=FALSE}
## PROBLEM 2

perm2 <- read.csv("PermTest2.csv")
set.seed(2000)

# observed difference in 90th percentiles between groups
p90_athlete <- quantile(na.omit(perm2$Athletes), probs = .90)
p90_nonathlete <- quantile(perm2$NonAthletes, probs = .90)

perc_obs <- p90_athlete - p90_nonathlete # 6.1

# specify lengths of groups
n1 <- length(na.omit(perm2$Athletes)); n2 <- length(perm2$NonAthletes)
N <- n1 + n2 # total observations

# comibine data
alldata <- c(na.omit(perm2$Athletes), perm2$NonAthletes)

# initialize variables
nperms = 1000
all_perc_perm <- c()
p_val2 <- 0
count2 <- 0

# permute data and calculate difference in 90th percentiles 1000 times
for(i in 1:nperms){ 
  # permute data
  index = sample(1:N, n1)
  data1 = alldata[index]
  data2 = alldata[-index]
  
  # calculate difference in 90th percentiles
  perc_a <- quantile(data1, probs = .90) # 90th percentile for athletes
  perc_na <- quantile(data2, probs = .90) # 90th percentile for non athletes
  perc_perm <- perc_a - perc_na
  all_perc_perm[i] <- perc_perm
  
  # for right-tailed p-value:
  # count2 = number of times the sample difference is >= the observed difference
  if(perc_perm > perc_obs | perc_perm == perc_obs){
    count2 <- count2 + 1
  }
  # calculate p-value
  p_val2 <- count2 / nperms
}

p_val2 # 0.04

# histrogram of permuted statistisc
ggplot() + 
  geom_histogram(aes(x=all_perc_perm), bins = 15, color="black", fill="plum2") +
  labs(x = "Difference in 90th Percentile Test Scores Between Athletes and Non-Athletes", 
       title="Distribution of 1000 Permuted Statistics", 
       y = "Count") +
  geom_vline(xintercept = perc_obs, linetype = "solid", color = "blue", size = .7) +
  geom_text(aes(x=5.1, label = "Observed", y=150), color="blue", angle=45) +
  theme_minimal()
```

**Ho:** The 90th percentile of test scores is the same between athletes and non-athletes (p_a - p_na = 0, where p_a is the population 90th percentile score for athletes)

**Ha:** The 90th percentile test score is greater for athletes than it is non-athletes (p_a - p_na \> 0)

The observed difference in 90th percentile test score between the athletes and non-athletes is 6.1. The p-value for the 1000 permutations is .04 \< .05, therefore, we would reject the null and have enough evidence to conclude that athletes have a greater 90th percentile score on standardized tests than non-athletes in the population.
